### 2015-05-18 papers/histogram/thesis/thesis.tex

\renewcommand{\title}{Optimizing Monte Carlo simulation of the
  square-well fluid}
\renewcommand{\author}{Michael A. Perlin}
\newcommand{\institution}{Oregon State University}
\newcommand{\department}{Department of Physics}
\newcommand{\supervisor}{Dr. David Roundy}
\renewcommand{\date}{08 May 2015}

@@

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Metropolis Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low (i.e. liquid-state) temperatures. Previous
works have developed several generic Monte Carlo histogram methods for
collecting statistics on low energy system states, but little or no
literature exists on these methods' systematic comparison, as well as
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model). We implement the Wang-Landau
method, transition matrix Monte Carlo, as well as two methods of our
own. The first of our methods, the simple flat method, is novel, while
the second is a hybrid of two published methods (transition matrix
Monte Carlo and the optimized ensemble).

We find that Wang-Landau potentially performs better than other
histogram methods, but fails catastrophically without a predefined
energy range for the square-well fluid. The simple flat method and
transition matrix Monte Carlo give results comparable to successful
Wang-Landau simulations, but simple flat has some anomalous cases with
large errors. Finally, our implementation of the optimized ensemble
using a transition matrix results in worse performance over the
straightforward transition matrix Monte Carlo method.

\fixme{edit when I'm awake}

@@

\fixme{give more intro about phase transitions and the critical
  point?}

Understanding the behavior of fluids near the critical point, or the
point at which the distinction between liquid and gas ceases to become
well-defined, is a major challenge in the field of classical density
functional theory. Studying the critical point of a fluid via
simulations requires exploring the boundary between its liquid and
gaseous phases. The square-well fluid is the simplest system with a
liquid-vapor phase transition, and is therefore of interest for
studying the critical point of fluids. For a generic square-well fluid
to be observed in a liquid state, it must have a localized
distribution of the spheres which make up the fluid. Such a
distribution implies a low energy fluid state, as all interactions
between spheres are attractive. Conversely, a low energy liquid state
of a square-well fluid otherwise observable in both liquid and gaseous
phases implies a localized distribution of spheres. The search for
liquid states of the square-well fluid is thus equivalent to the
search for low energy states.

Monte Carlo simulations are a standard means for studying the
equilibrium thermodynamic properties of a system. In the most
straightforward implementation of Monte Carlo fluid simulations,
however, highly localized distributions of fluid elements are
extremely unlikely to occur, making such simulations impractical for
studying the critical point of the square-well fluid. Several generic
computational methods, called histogram methods, exist for dealing
with such a problem; namely, that interesting regions of a system's
state space cannot be sampled via standard Monte Carlo methods in a
reasonable amount of time. There is no common consensus, however, on
the relative efficacy of these methods. Furthermore, some methods,
such as the Wang-Landau method\cite{wang_landau}, require tuning
several free parameters that lack both ``obvious'' values and
systematic means for assigning them. This work will thus attempt to
quantify and compare the effectiveness of a few such methods as
applied to the square-well fluid.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.47\textwidth]{figs/square-well.pdf}
  \caption[The square-well pair potential]
  {The square-well pair potential can be used to model short range
    forces to first order. The infinite potential at $r<\sigma$ simply
    enforces the condition that spheres cannot overlap. At
    $r>\lambda\sigma$, the potential is null, and makes no
    contribution to the fluid's internal energy.}
  \label{fig:pair_potential}
\end{figure}

@@

\subsection{Monte Carlo fluid simulation}
\label{sec:mc_sim}

Model systems such as the square-well fluid are powerful tools for
understanding complex physical systems, but they do not themselves
exist in the real world. Consequently, direct experimental tests of
theories for model system (e.g. square-well density functional
theories) are not possible. For this reason, model system theories are
commonly tested against Monte Carlo simulations. Proper implementation
of Monte Carlo methods to study completely characterized systems
ensures that statistical results from simulations converge on the
exact properties of the simulated system in the infinite simulation
time limit. Furthermore, uncertainties in quantities computed via
Monte Carlo simulations are typically well-defined, monotonically
decreasing functions of simulation time, allowing one to run
simulations to the desired level of accuracy.

@@

\begin{algorithm}[tb]
  \caption{Metropolis Monte Carlo fluid simulation}
  \label{alg:metropolis}
  \begin{alg}

  \item Construct an initial ``typical'' (i.e. non-ordered) fluid
    configuration. \fixme{explain how}

@@

Algorithm \ref{alg:metropolis} provides a sketch of unbiased,
Metropolis Monte Carlo fluid simulation. Such an algorithm is
``unbiased'' in the sense that it collects statistics (i.e. data) on
all valid system configurations with equal probability. Statistics
whose collection time scales as $\mathcal O\p{1}$ with system size,
meaning that increasing the number of spheres $N$ in the simulation
does not affect collection time, can be collected after every move
(defined in the algorithm). Statistics whose collection time scales as
$\mathcal O\p{N}$, meaning that doubling $N$ doubles collection time,
can be collected after each iteration. In general, collection with
$\mathcal O\sp{\chi\p{N}}$ time scaling should not occur more often
than once every $\chi\p{N}$ moves, where $\chi\p{N}$ may be an
arbitrary function of $N$ void of constant prefactors, e.g. $N\log N$,
or $2^N$. These collection rules ensure that scaling up simulations
does not cause them to asymptotically spend all computation time only
collecting statistics, or only simulating the fluid. Collected
statistics are used to find thermodynamic properties of the simulated
fluid.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/dos-example.pdf}
  \caption[Density of states]
  {Sample density of states for square-well fluid with a well width
    $\lambda=1.3$, filling fraction $\eta=0.1$, and $N=100$ spheres,
    computed via the transition matrix Monte Carlo method. The density
    of states has been normalized to have a maximum value of $1$.}
  \label{fig:dos_sample}
\end{figure}

@@

The partition function at $T=T_0$ for simulations with canonical
weights is
\begin{align}
  \tilde Z\p{T_0}=\sum_E\tilde D\p{E}e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{w\p{E}}~e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{e^{-E/kT_0}}~e^{-E/kT_0}=\sum_EH\p{E},
  \label{Z_canonical}
\end{align}
and the value of a thermodynamic property $\bk{X}_{T_0}$
\begin{align}
  \bk{X}_{T_0}=\f1{\tilde Z\p{T_0}}\sum_E\bk{X}_E\tilde
  D\p{E}e^{-E/kT_0} =\f{\sum_E\bk{X}_EH\p{E}}{\sum_EH\p{E}},
  \label{eq:X_canonical}
\end{align}
where $\bk{X}_E$ is the mean value of $X$ at the energy $E$. The
simplifications in in (\ref{Z_canonical}) and (\ref{eq:X_canonical}),
which have no explicit dependence on $T_0$, occur because canonical
Monte Carlo simulations sample energies in proportion to the
distribution (over energy) of microstates at a temperature of $T_0$.
Canonical Monte Carlo simulations can therefore fail to sufficiently
sample energies which are important (i.e. energies with a
non-negligible state probability density) at different
temperatures. As a consequence, such simulations should not be used to
determine properties $\bk{X}_{T\ne T_0}$, and are thus referred to as
``fixed temperature'' simulations. Incidentally, Metropolis Monte
Carlo simulations can be thought of as infinite-temperature
simulations, as they are equivalent to simulations with canonical
weights
\begin{align}
  w\p{E}=\lim_{T\to\infty}e^{-E/kT}=1.
  \label{eq:inf_temp_weights}
\end{align}
This observation will be used in Section \ref{sec:min_energy} to
identify a maximum energy of interest.

@@

The Wang-Landau method is the first published histogram method we
discuss\cite{wang_landau, wang_landau_mod, wang_landau_analysis}. This
method has been used to study a wide variety of systems\fixme{which
  systems? cite}, and is the standard method\fixme{cite} to achieve
broad energy sampling via Monte Carlo simulations. \fixme{more intro
  on WL}

@@

While one could imagine many initialization routines to determine the
transition matrix $T$, we use Algorithm \ref{alg:tmmc}, which is a
modified version of a routine provided in \cite{tmmc_prl}. \fixme{is
  this citation correct?} The primary motivation behind this algorithm
is such: as lower energies are always more difficult to sample than
higher energies, transitions to lower energies should always be
accepted, while transitions to higher energies should pass a
probabilistic acceptance test. To understand the probability in step
\ref{alg:tmmc_prob} of Algorithm \ref{alg:tmmc} of accepting a move
$s_i\to s_f$ for which $E_f>E_i$, first consider the case of $c=0$ and
$T_{\t{min}}\to0$, in which case the probability becomes
\begin{align}
  P_m\p{s_i\to s_f}=\f{\p{\tilde T_d\p{E_i,\Delta E}}\big/
    \p{\sum_{\Delta E'}\tilde T_d\p{E_i,\Delta E'}}} {\p{\tilde
      T_d\p{E_f,-\Delta E}}\big/ \p{\sum_{\Delta E'}\tilde
      T_d\p{E_f,\Delta E'}}} =\f{T_d\p{E_i,\Delta
      E}}{T_d\p{E_f,-\Delta E}} =\f{P\p{E_i\to E_f}}{P\p{E_f\to E_i}},
  \label{eq:tmmc_prob_simplified}
\end{align}
where $\Delta E=E_f-E_i$ and $P\p{E_m\to E_n}$ is the unbiased
probability of a transition from an energy $E_m$ to an energy
$E_n$. The result in (\ref{eq:tmmc_prob_simplified}) is the
probability of accepting moves $s_i\to s_f$ required for simulation
via Algorithm \ref{alg:tmmc} to yield a flat energy
histogram\cite{tmmc}. Unlike the previous histogram methods, we desire
a flat histogram during initialization via Algorithm \ref{alg:tmmc}
not because we are constructing weights $w\p{E}$, but because we wish
to sample the transition histogram $\tilde T_d\p{E,\Delta E}$ equally
at all energies $E$. \fixme{briefly explain
  (\ref{eq:tmmc_prob_simplified}) by considering a two-level system,
  and extending to larger systems}.

@@

The hybrid OE-TMMC method thus first runs Algorithm \ref{alg:tmmc} to
collect a transition histogram and compute flat histogram weights,
then computes the diffusivity $\alpha\p{E}$ via
(\ref{eq:<de>})-(\ref{eq:microscopic_diffusitivy}), and finally
computes optimized weights as given by
(\ref{eq:optimized_weights}). Unlike the optimized ensemble, OE-TMMC
does not require any extra simulation routines, and is therefore
straightforward to implement and compare to other histogram methods.
\fixme{provide iterative algorithm?}

@@

Our analysis will involve comparing errors in system properties
computed from simulation results. As we do not know the true values of
these system properties, we will compare standard simulation results
against those from ``gold standard'' simulations, which run for
significantly longer than the standard $10^5N^2$ iterations. For our
gold standard simulations, we use the TMMC method and initialize until
it collects $10^4$ optimistic samples of the minimum important energy
(i.e. until $s_o\p{E_-}\ge 10^4$). \fixme{why TMMC?} To increase
confidence in low-temperature gold standard results, we initialized
them with $kT_{\t{min}}=0.1\epsilon$. After initialization, the gold
standard runs simulate for $10^{10}$ iterations ($10^{10}N$
moves). While the entire initialization and simulation process takes
around two hours for the largest regular simulations, the largest
golden simulations initialize for $\sim11$ hours, and simulate for a
week. To control for iteration number in the quality of results, we
compare against ``converged flat'' simulations. Converged flat
simulations read in the transition matrix from golden simulations to
construct a weight array, and simulate for the same number of
iterations as the standard simulations.

@@

The most immediate observation in these plots is the failure of
Wang-Landau to produce meaningful results without a predefined energy
range. This failure occurs because the Wang-Landau algorithm does not
have a mechanism for properly setting and adjusting weights on newly
observed energies. After several reset cycles of the Wang-Landau
algorithm, the weight-modifying factor $f_k$ gets exceedingly small
relative to its initial value. When Wang-Landau finds new low
energies, therefore, the simulation gets stuck trying to fix the
weights at these energies with a factor $f_k$ whose magnitude is
appropriate for fine tuning, rather than heavy adjustment. If
Wang-Landau knows the energy range of interest to begin with, it will
not decrease $f_k$ until it has observed all important energies, and
adjusted their weights to some reasonable value. Due to the drastic
failure of Wang-Landau, will omit its nonsensical results from some
figures. \fixme{which figures?}

@@

High energy histogram counts at low energies mean that a simulation
has collected many statistics on these energies, but the energy
histogram contains no information about how correlated these
statistics are. To determine how effective different simulations are
at collecting uncorrelated statistics, we can look at the pessimistic
sample rate $r_p\p{E}$, which is the mean number of simulation
iterations required to sample a given energy $E$. This rate is roughly
equal to the number of iterations $I$ in a simulation and the
pessimistic sample rate $s_p\p{E}$ by $r_p\p{E}=I/s_p\p{E}$; the
greater the number of iterations $I$, the better this
approximation. Figure \ref{fig:sample_rates} shows pessimistic sample
rates averaged over 30 simulations with $N=25$. Lower pessimistic
sample rates mean that a simulation takes less time to collect
statistics which are guaranteed to be uncorrelated. By this metric,
TMMC and OE-TMMC again perform worse than simple flat and vanilla
Wang-Landau.\fixme{comment on why it is interesting that OETMMC doing
  worse than TMMC?}

@@

Though Figures \ref{fig:internal_energy}--\ref{fig:config_entropy} are
interesting for studying the square-well fluid, they do not help us
resolve the efficacy of different histogram methods. These figures do,
however, show that the square-well fluid has a phase transition around
$kT\approx0.5\epsilon$, which justifies our choice of
$kT_{\t{min}}=0.2\epsilon$ and ensures that the histogram methods are
indeed performing the nontrivial task of exploring the liquid-gas
phase boundary of the square-well fluid. For the purposes of this
paper, we will neglect to demonstrate that this phase transition is
indeed between liquid and gas phases, and take this fact as
given. \fixme{cite instead of stating as fact?}

@@

After implementing several histogram methods and applying them to the
$\lambda=1.3$, $\eta=0.3$ square-well fluid, we found that
Wang-Landau, if given a predefined energy range, consistently yields
some of the smallest errors and highest low-energy sampling rates of
all histogram methods, with comparable results from TMMC and the
simple flat method. Without prior knowledge of the minimum important
energy of a system, however, Wang-Landau can get stuck at low
energies, and is incapable of recovering. Wang-Landau is therefore
most appropriate for studying systems in which the entire energy range
is known a priori, or in which only a given energy range is of
particular interest. The simple flat method also sometimes fails to
initialize; if gone unnoticed, such failures can be costly in terms of
computation time and resources. Our implementation of the optimized
ensemble using the transition matrix resulted in lower sampling rates
and higher errors than the flat histogram TMMC weights. We suspect
that the optimized ensemble is failing to improve TMMC because its
theory neglects the discretization of energies, which may be
non-negligible for the systems we simulated.

\fixme{make the conclusion more satisfying?}

>>>>

\renewcommand{\title}{Optimizing Monte Carlo simulation of the
  square-well fluid}
\renewcommand{\author}{Michael A. Perlin}
\newcommand{\institution}{Oregon State University}
\newcommand{\department}{Department of Physics}
\newcommand{\supervisor}{Dr. David Roundy}
\renewcommand{\date}{15 May 2015}

@@

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Metropolis Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low liquid-state temperatures. Previous works
have developed several generic Monte Carlo histogram methods for
collecting statistics on low energy system states, but little or no
literature exists on these methods' systematic comparison, as well as
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model).

We implement our own ``simple flat'' method, the Wang-Landau method,
transition matrix Monte Carlo, and a modified version of the optimized
ensemble. Performance of each method is determined by low energy
sampling rates and maximum errors in several computed system
properties. We find that Wang-Landau potentially performs better than
other histogram methods, but fails catastrophically without a
predetermined energy range. The simple flat method and transition
matrix Monte Carlo give results comparable to successful Wang-Landau
simulations, but simple flat has some anomalous cases with large
errors. Finally, our implementation of the optimized ensemble using a
transition matrix results in worse performance over the
straightforward transition matrix Monte Carlo method.

\fixme{edit}

@@

The critical point of a system is the point in phase space at which
the distinction between liquid and gaseous phases ceases to be
well-defined. As shown in Figure \ref{fig:phase_diagram}, a fluid need
not boil or condense to transition between liquid to gaseous phases:
it can smoothly transition between the two by circumventing the
critical point. Understanding the behavior of fluids near the critical
point is a major challenge in the field of classical density
functional theory. The square-well fluid is the simplest system with a
liquid-vapor phase transition, and is therefore of interest for
studying the critical point. For a generic square-well fluid to be
observed in a liquid state, it must have a localized distribution of
the spheres which make up the fluid. Such a distribution implies a low
energy fluid state, as all interactions between spheres are
attractive. Conversely, a low energy liquid state of a square-well
fluid otherwise observable in both liquid and gaseous phases implies a
localized distribution of spheres. The search for liquid states of the
square-well fluid is thus equivalent to the search for low energy
states.

Monte Carlo simulations are a standard means for studying the
equilibrium thermodynamic properties of a system. In the most
straightforward implementation of Monte Carlo fluid simulations,
however, highly localized distributions of spheres are extremely
unlikely to occur, making such simulations impractical for studying
the critical point of the square-well fluid. Several generic
computational methods, called histogram methods, exist for dealing
with such a problem; namely, that interesting regions of a system's
state space cannot be sampled via standard Monte Carlo methods in a
reasonable amount of time. There is little consensus, however, on the
relative efficacy of these methods, and few have applied them to
simulation of the square-well fluid.

In this paper, we first review what the square-well fluid is, as well
as how to simulate it via standard (Metropolis) Monte Carlo (Sections
\ref{sec:sw_fluid}--\ref{sec:monte_carlo}). We then address the
shortcomings of this means to study the square-well fluid, and
consider how one can modify simulations to overcome these shortcomings
in Section \ref{sec:histogram_methods}. We develop a thorough
understanding of several histogram methods in Sections
\ref{sec:flat_histogram}--\ref{sec:oetmmc}, and discuss some details
pertinent to the actual implementation of these methods in Section
\ref{sec:loose_ends}. Finally, in Section \ref{sec:results} we analyze
the performance of a few histogram methods in simulations of a
particular square-well fluid.

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.47\textwidth]{figs/phase-diagram.pdf}
  \caption[Phase diagram]
  {A generic phase diagram showing the boundaries between solid,
    liquid, and gaseous phases.}
  \label{fig:phase_diagram}
\end{figure}

@@

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.47\textwidth]{figs/square-well.pdf}
  \caption[The square-well pair potential]
  {The square-well pair potential can be used to model short range
    forces to first order. The infinite potential at $r<\sigma$ simply
    enforces the condition that spheres cannot overlap. At
    $r>\lambda\sigma$, the potential is null, and makes no
    contribution to the fluid's internal energy.}
  \label{fig:pair_potential}
\end{figure}

\subsection{Monte Carlo fluid simulation}
\label{sec:monte_carlo}

Model systems, such as the square-well fluid, are powerful tools for
understanding complex physical systems, but they do not themselves
exist in the real world. Consequently, direct experimental tests of
theories for model system (e.g. square-well density functional
theories) are not possible. For this reason, model system theories are
commonly tested against Monte Carlo simulations. Proper implementation
of Monte Carlo methods to study completely characterized systems
ensures that statistical results from simulations converge on the
exact properties of the simulated system in the infinite simulation
time limit. Furthermore, uncertainties in quantities computed via
Monte Carlo simulations are typically well-defined, monotonically
decreasing functions of simulation time, allowing one to run
simulations to the desired level of accuracy.

@@

\begin{algorithm}[!b]
  \caption{Metropolis Monte Carlo fluid simulation}
  \label{alg:metropolis}
  \begin{alg}

  \item Construct an initial ``typical'' (i.e. non-ordered) fluid
    configuration. \fixme{explain how?}

@@

Algorithm \ref{alg:metropolis} provides a sketch of unbiased,
Metropolis Monte Carlo fluid simulation. Such an algorithm is
``unbiased'' in the sense that it collects statistics (i.e. data) on
all valid system configurations with equal probability. Statistics
whose collection time scales as $\mathcal O\p{1}$ with system size,
meaning that increasing the number of spheres $N$ in the simulation
does not affect collection time, can be collected after every move
(defined in the algorithm). Statistics whose collection time scales as
$\mathcal O\p{N}$, meaning that doubling $N$ doubles collection time,
can be collected after each iteration. In general, collection with
$\mathcal O\sp{\chi\p{N}}$ time scaling should not occur more often
than once every $\chi\p{N}$ moves, where $\chi\p{N}$ may be an
arbitrary function of $N$, e.g. $N\log N$, or $2^N$. These collection
rules ensure that scaling up simulations does not cause them to
asymptotically spend all computation time only collecting statistics,
or only simulating the fluid. Collected statistics are used to find
thermodynamic properties of the simulated fluid.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/dos-thesis-example.pdf}
  \caption[Density of states]
  {Sample density of states for square-well fluid with a well width
    $\lambda=1.3$, filling fraction $\eta=0.1$, and $N=100$ spheres,
    computed via the transition matrix Monte Carlo method. The density
    of states has been normalized to have a maximum value of $1$.}
  \label{fig:dos_sample}
\end{figure}

@@

The partition function at $T=T_0$ for simulations with canonical
weights is
\begin{align}
  \tilde Z\p{T_0}=\sum_E\tilde D\p{E}e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{w\p{E}}~e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{e^{-E/kT_0}}~e^{-E/kT_0}=\sum_EH\p{E},
  \label{Z_canonical}
\end{align}
and the value of a thermodynamic property $\bk{X}_{T_0}$
\begin{align}
  \bk{X}_{T_0}=\f1{\tilde Z\p{T_0}}\sum_E\bk{X}_E\tilde
  D\p{E}e^{-E/kT_0} =\f{\sum_E\bk{X}_EH\p{E}}{\sum_EH\p{E}},
  \label{eq:X_canonical}
\end{align}
where $\bk{X}_E$ is the mean value of $X$ at the energy $E$. The
simplifications in in (\ref{Z_canonical}) and (\ref{eq:X_canonical}),
which have no explicit dependence on $T_0$, occur because canonical
Monte Carlo simulations sample energies in proportion to the
distribution (over energy) of microstates at a temperature of $T_0$.
Canonical Monte Carlo simulations can therefore fail to sufficiently
sample energies which are important (i.e. energies with a
non-negligible state probability density) at different
temperatures. As a consequence, such simulations should not be used to
determine properties $\bk{X}_{T\ne T_0}$, and are thus referred to as
``fixed temperature'' simulations. Incidentally, Metropolis Monte
Carlo simulations can be thought of as infinite-temperature
simulations, as they are equivalent to simulations with canonical
weights
\begin{align}
  w\p{E}=\lim_{T\to\infty}e^{-E/kT}=1.
  \label{eq:inf_temp_weights}
\end{align}
This observation will be used in Section \ref{sec:min_energy} to
identify a maximum energy of interest in simulations.

@@

The Wang-Landau method is the first published histogram method we
discuss\cite{wang_landau, wang_landau_mod, wang_landau_analysis}. This
method has been used to study a wide variety of systems\fixme{which
  systems? cite}, and is the standard method\fixme{cite} to achieve
broad energy sampling via Monte Carlo simulations.

@@

While one could imagine many initialization routines to determine the
transition matrix $T$, we use Algorithm \ref{alg:tmmc}, which is a
modified version of a routine provided in \cite{tmmc_prl}. \fixme{is
  this citation correct?} The primary motivation behind this algorithm
is such: as lower energies are always more difficult to sample than
higher energies, transitions to lower energies should always be
accepted, while transitions to higher energies should pass a
probabilistic acceptance test. To understand the probability in step
\ref{alg:tmmc_prob} of Algorithm \ref{alg:tmmc} of accepting a move
$s_i\to s_f$ for which $E_f>E_i$, first consider the case of $c=0$ and
$T_{\t{min}}\to0$, in which case the probability becomes
\begin{align}
  P_m\p{s_i\to s_f}=\f{\p{\tilde T_d\p{E_i,\Delta E}}\big/
    \p{\sum_{\Delta E'}\tilde T_d\p{E_i,\Delta E'}}} {\p{\tilde
      T_d\p{E_f,-\Delta E}}\big/ \p{\sum_{\Delta E'}\tilde
      T_d\p{E_f,\Delta E'}}} =\f{T_d\p{E_i,\Delta
      E}}{T_d\p{E_f,-\Delta E}} =\f{P\p{E_i\to E_f}}{P\p{E_f\to E_i}},
  \label{eq:tmmc_prob_simplified}
\end{align}
where $\Delta E=E_f-E_i$ and $P\p{E_m\to E_n}$ is the unbiased
probability of a transition from an energy $E_m$ to an energy
$E_n$. The result in (\ref{eq:tmmc_prob_simplified}) is the
probability of accepting moves $s_i\to s_f$ required for simulation
via Algorithm \ref{alg:tmmc} to yield a flat energy
histogram\cite{tmmc}. Unlike the previous histogram methods, we desire
a flat histogram during initialization via Algorithm \ref{alg:tmmc} in
order to broadly sample the transition histogram $\tilde
T_d\p{E,\Delta E}$, rather than to construct weights $w\p{E}$
directly.

@@

The hybrid OE-TMMC method thus first runs Algorithm \ref{alg:tmmc} to
collect a transition histogram and compute flat histogram weights,
then computes the diffusivity $\alpha\p{E}$ via
(\ref{eq:<de>})-(\ref{eq:microscopic_diffusitivy}), and finally
computes optimized weights as given by
(\ref{eq:optimized_weights}). Unlike the optimized ensemble, OE-TMMC
does not require any extra simulation routines, and is therefore
straightforward to implement and compare to other histogram methods.

@@

Our analysis will involve comparing errors in system properties
computed from simulation results. As we do not know the true values of
these system properties, we will compare standard simulation results
against those from ``gold standard'' simulations, which run for
significantly longer than the standard $10^5N^2$ iterations. For our
gold standard simulations, we use the TMMC method and initialize until
it collects $10^4$ optimistic samples of the minimum important energy
(i.e. until $s_o\p{E_-}\ge 10^4$). \fixme{why TMMC?} To increase
confidence in low-temperature gold standard results, we initialized
them with $kT_{\t{min}}=0.1\epsilon$. After initialization, the gold
standard runs simulate for $10^{10}$ iterations ($10^{10}N$
moves). While the entire initialization and simulation process takes
around two hours for the largest regular simulations, the largest
golden simulations initialize for $\sim11$ hours, and simulate for a
week. To control for iteration number in the results from any given
set of simulations, we compare against ``converged flat''
simulations. Converged flat simulations read in the transition matrix
from golden simulations to construct a weight array, and simulate for
the same number of iterations as the standard simulations.

@@

The most immediate observation in these plots is the failure of
Wang-Landau to produce meaningful results without a predefined energy
range. This failure occurs because the Wang-Landau algorithm does not
have a mechanism for properly setting and adjusting weights on newly
observed energies. After several reset cycles of the Wang-Landau
algorithm, the weight-modifying factor $f_k$ gets exceedingly small
relative to its initial value. When Wang-Landau finds new low
energies, therefore, the simulation gets stuck trying to fix the
weights at these energies with a factor $f_k$ whose magnitude is
appropriate for fine tuning, rather than heavy adjustment. If
Wang-Landau knows the energy range of interest to begin with, it will
not decrease $f_k$ until it has observed all important energies, and
adjusted their weights to some reasonable value. Due to the drastic
failure of Wang-Landau, will omit its nonsensical results from some
figures, as these results are distracting and unfit for drawing any
conclusions.

@@

High energy histogram counts at low energies mean that a simulation
has collected many statistics on these energies, but the energy
histogram contains no information about how correlated these
statistics are. To determine how effective different simulations are
at collecting uncorrelated statistics, we can look at the pessimistic
sample rate $r_p\p{E}$, which is the mean number of simulation
iterations required to sample a given energy $E$. This rate is roughly
equal to the number of iterations $I$ in a simulation and the
pessimistic sample rate $s_p\p{E}$ by $r_p\p{E}=I/s_p\p{E}$; the
greater the number of iterations $I$, the better this
approximation. Figure \ref{fig:sample_rates} shows pessimistic sample
rates averaged over 30 simulations with $N=25$. Lower pessimistic
sample rates mean that a simulation takes less time to collect
statistics which are guaranteed to be uncorrelated. By this metric,
TMMC and OE-TMMC again perform worse than simple flat and vanilla
Wang-Landau.

@@

Though Figures \ref{fig:internal_energy}--\ref{fig:config_entropy} are
interesting for studying the square-well fluid, they do not help us
resolve the efficacy of different histogram methods. These figures do,
however, show that the square-well fluid has a phase transition around
$kT\approx0.5\epsilon$, which justifies our choice of
$kT_{\t{min}}=0.2\epsilon$ and ensures that the histogram methods are
indeed performing the nontrivial task of exploring the liquid-gas
phase boundary of the square-well fluid. For the purposes of this
paper, we will neglect to prove that this phase transition is indeed
between liquid and gas phases, and take this fact as given. \fixme{say
  more convincingly}

@@

After implementing several histogram methods and applying them to
simulation of the $\lambda=1.3$, $\eta=0.3$ square-well fluid, we
found that Wang-Landau, if given a predefined energy range,
consistently yields some of the smallest errors and highest low-energy
sampling rates of all histogram methods, with comparable results from
TMMC and the simple flat method. Without prior knowledge of the
minimum important energy of a system, however, Wang-Landau can get
stuck at low energies, in which case it is incapable of
recovering. Wang-Landau is therefore most appropriate for studying
systems in which the entire energy range is known a priori, or in
which only a given energy range is of particular interest. The simple
flat method, while generally performing well, sometimes fails to
initialize. If gone unnoticed, such failures can be costly in terms of
computation time and resources. Our implementation of the optimized
ensemble using the transition matrix resulted in lower sampling rates
and higher errors than the flat histogram TMMC weights. We suspect
that the optimized ensemble is failing to improve TMMC because the
theory behind it neglects the discretization of energies.

\fixme{make conclusions more satisfying?}
